---
title: "Morgan Stanley Employee Sentiment Analysis"
subtitle: "Web Scraping & NLP with Glassdoor and Indeed Reviews"
author: "Slobodan"
date: "Original: 01.09.2024., edited for portfolio: 27.03.2026."
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  eval = TRUE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)
```

---

## Overview

This project explores employee sentiment at **Morgan Stanley** using reviews scraped from **Glassdoor** and **Indeed**. The full pipeline covers:

1. **Web scraping** both platforms via the Scrapfly API
2. **Data wrangling** - cleaning, combining, and tokenising the reviews
3. **Sentiment analysis** using three NLP lexicons (NRC, Bing, and AFINN)
4. **Temporal analysis** - tracking how sentiment has shifted across four time periods (Pre-COVID, Post-COVID, 7–12 months ago, and the Last 6 months (February 2024 - September 2024)

The analysis focuses on **former employees**, whose reviews tend to be more candid.

---

## Part 1 - Glassdoor Scraper

### Libraries & API Configuration

The scraper relies on [`crul`](https://docs.ropensci.org/crul/) for HTTP requests and routes them through the [Scrapfly](https://scrapfly.io/) proxy API to bypass anti-scraping protections.

```{r glassdoor-libs, eval=FALSE}
library(crul)
library(stringr)
library(jsonlite)
library(dplyr)
library(parallel)

SCRAPFLY_KEY <- ""
```

### Scrapfly Helper Function

This helper builds the proxied request URL. It supports JavaScript rendering, country-level proxy selection, and anti-scraping protection (`asp = TRUE`).

```{r glassdoor-helper, eval=FALSE}
scrapfly <- function(url,
                     render_js       = FALSE,
                     wait_for_selector = NULL,
                     asp             = FALSE,
                     country         = "us",
                     proxy_pool      = "public_datacenter_pool") {
  url_build("https://api.scrapfly.io/scrape", query = list(
    key               = SCRAPFLY_KEY,
    url               = url,
    asp               = "true",
    render_js         = if (render_js) "true" else "false",
    wait_for_selector = wait_for_selector,
    country           = country,
    proxy_pool        = proxy_pool
  ))
}
```

### Single-Page Scraper

Each Glassdoor review page is scraped individually. The function parses the JSON response and extracts review fields using regex. It retries up to 5 times on failure and adds a short delay between attempts.

```{r glassdoor-single, eval=FALSE}
scrape_single_page <- function(url) {
  retry  <- 1
  success <- FALSE
  reviews_df <- data.frame()

  while (!success && retry <= 5) {
    tryCatch({
      response     <- HttpClient$new(scrapfly(url))$get()
      data         <- jsonlite::fromJSON(response$parse())
      html_string  <- data$result$content
      cleaned_content <- gsub("\n", "", html_string)

      # Initialise storage vectors
      dates <- current <- urls <- titles <- pros <- cons <- vector("character", 0)

      # Split on review boundaries and drop trailing element
      reviews <- strsplit(cleaned_content, ',"summaryOriginal":', fixed = TRUE)[[1]]
      reviews <- reviews[-length(reviews)]

      # Extract fields with regex
      full    <- reviews
      dates   <- str_match(reviews, '"reviewDateTime":"(.*?)"')[, 2]
      current <- str_match(reviews, '"isCurrentJob":(true|false)')[, 2]
      urls    <- str_match(reviews, '"reviewId":(\\d+)')[, 2]
      titles  <- str_match(reviews, '"summary":"(.*?)"')[, 2]
      pros    <- str_match(reviews, '"pros":"(.*?)"')[, 2]
      cons    <- str_match(reviews, '"cons":"(.*?)"')[, 2]

      reviews_df <- data.frame(
        Full_review        = full,
        Date               = dates,
        IsCurrentEmployee  = current,
        URL                = urls,
        Title              = titles,
        Pros               = pros,
        Cons               = cons,
        stringsAsFactors   = FALSE
      )

      # Append reviewer position
      positions <- str_match_all(
        cleaned_content,
        '"author":\\{"@type":"Person","name":"(.*?)"\\}'
      )[[1]][, 2]
      reviews_df$Position <- positions

      success <- TRUE

    }, error = function(e) {
      cat("Error on page - retrying (attempt", retry, "of 5)\n")
      retry <<- retry + 1
      Sys.sleep(5)
    })
  }

  if (!success) cat("Failed after 5 attempts. Skipping page...\n")
  return(reviews_df)
}
```

### Multi-Page Scraper

Pages are scraped sequentially, with progress auto-saved every 30 pages as `.rds` checkpoints to protect against interruptions.

```{r glassdoor-multi, eval=FALSE}
scrape_all_pages <- function(base_url, max_pages, save_interval = 30) {
  all_reviews <- data.frame()

  for (i in 1:max_pages) {
    url <- if (i == 1) {
      base_url
    } else {
      gsub("_IP1", paste0("_IP", i), base_url)
    }

    cat("Scraping page", i, "of", max_pages, "...\n")
    page_reviews <- scrape_single_page(url)
    all_reviews  <- rbind(all_reviews, page_reviews)

    if (i %% save_interval == 0 || i == max_pages) {
      save_file <- paste0("glassdoor_reviews_page_", i, ".rds")
      saveRDS(all_reviews, file = save_file)
      cat("Saved progress at page", i, "→", save_file, "\n")
    }

    Sys.sleep(2)
  }

  cat("Scraping complete. Last page:", i, "\n")
  return(all_reviews)
}
```

### Execution & Export

```{r glassdoor-run, eval=FALSE}
base_url <- paste0(
  "https://www.glassdoor.com/Reviews/",
  "Morgan-Stanley-US-Reviews-EI_IE2282.0,14_IL.15,17_IN1_IP1.htm",
  "?filter.iso3Language=eng",
  "&filter.employmentStatus=REGULAR",
  "&filter.employmentStatus=PART_TIME"
)
max_pages <- 616

all_reviews_df <- scrape_all_pages(base_url, max_pages)

# Drop the raw Full_review column before export
reviews_no_full <- all_reviews_df[-1]
write.csv(reviews_no_full, "Glassdoor Morgan Stanley reviews.csv", row.names = FALSE)
```

---

## Part 2 - Indeed Scraper


> The `scrapfly()` helper is identical to the one defined in Part 1.

```{r indeed-helper, eval=FALSE}
scrapfly <- function(url,
                     render_js         = FALSE,
                     wait_for_selector = NULL,
                     asp               = FALSE,
                     country           = "us",
                     proxy_pool        = "public_datacenter_pool") {
  url_build("https://api.scrapfly.io/scrape", query = list(
    key               = SCRAPFLY_KEY,
    url               = url,
    asp               = "true",
    render_js         = if (render_js) "true" else "false",
    wait_for_selector = wait_for_selector,
    country           = country,
    proxy_pool        = proxy_pool
  ))
}
```

### Page Parser

Indeed embeds review data as JSON within the HTML. The parser isolates the `reviewsList` block by finding its start and end markers before extracting individual fields.

```{r indeed-parser, eval=FALSE}
get_review_page <- function(url) {
  response    <- HttpClient$new(scrapfly(url))$get()
  data        <- jsonlite::fromJSON(response$parse())
  html_string <- data$result$content

  # Isolate the reviews JSON block
  start_marker <- '"reviewsList":'
  end_marker   <- '"paginationLinks":'
  start_pos    <- regexpr(start_marker, html_string)
  end_pos      <- regexpr(end_marker,   html_string)

  reviews_section <- if (start_pos != -1 & end_pos != -1) {
    substr(html_string, start_pos, end_pos)
  } else {
    NA
  }

  # Extract each field
  positions <- str_match_all(reviews_section, '"jobTitle":"(.*?)",' )[[1]][, 2]
  positions <- positions[positions != ""]
  dates     <- str_match_all(reviews_section, '"submissionDate":"(.*?)"')[[1]][, 2]
  urls      <- str_match_all(reviews_section, '"reviewUrl":"(.*?)"')[[1]][, 2]
  reviews   <- str_match_all(reviews_section, '"text":\\{"text":"(.*?)"\\}')[[1]][, 2]
  titles    <- str_match_all(reviews_section, '"title":\\{"text":"(.*?)"\\}')[[1]][, 2]
  current   <- str_match_all(reviews_section, '"currentEmployee":(true|false)')[[1]][, 2]
  location  <- str_match_all(reviews_section, '"location":"(.*?)"')[[1]][, 2]

  data.frame(
    Full_page        = reviews_section,
    Date             = dates,
    Position         = positions,
    Current_employee = current,
    URL              = urls,
    Title            = titles,
    Review           = reviews,
    Location         = location,
    stringsAsFactors = FALSE
  )
}
```

### Execution & Export

Indeed paginates in steps of 20. Page 0 is scraped first, followed by pages 20–400 in a loop.

```{r indeed-run, eval=FALSE}
all_reviews <- data.frame(
  Full_page        = character(),
  Date             = character(),
  Position         = character(),
  Current_employee = character(),
  URL              = character(),
  Title            = character(),
  Review           = character(),
  Location         = character(),
  stringsAsFactors = FALSE
)

# First page (offset 0)
url          <- paste0('https://www.indeed.com/cmp/Mckinsey-&-Company/reviews?start=0&lang=en')
page_reviews <- get_review_page(url)
all_reviews  <- bind_rows(all_reviews, page_reviews)

# Remaining pages
for (i in seq(20, 400, by = 20)) {
  url          <- paste0('https://www.indeed.com/cmp/Mckinsey-&-Company/reviews?start=', i, '&lang=en')
  cat("Scraping offset:", i, "\n")
  page_reviews <- get_review_page(url)
  all_reviews  <- bind_rows(all_reviews, page_reviews)
  Sys.sleep(1)
}

write.csv(all_reviews, "Indeed_reviews.csv", row.names = FALSE)
```

---

## Part 3 - Sentiment Analysis

### Libraries

```{r sa-libs}
library(tidyverse)
library(readr)
library(tidytext)
library(forcats)
```

### Load & Prepare Data

The two datasets are loaded, harmonised to a shared column schema, and merged into a single dataframe.

```{r sa-load, eval=TRUE}
Glassdoor <- read_csv("Glassdoor Morgan Stanley reviews.csv")
Indeed    <- read_csv("Indeed Morgan Stanley reviews.csv")
```

```{r sa-wrangle, eval=TRUE}
# Glassdoor
Glassdoor <- Glassdoor %>%
  mutate(Review = paste(Pros, Cons, sep = " ")) %>%   # combine Pros + Cons into one Review field
  rename(Current_employee = IsCurrentEmployee) %>%
  mutate(
    Date     = ymd_hms(Date),
    Platform = "Glassdoor"
  ) %>%
  rename(ID = ...1)

# Indeed
Indeed <- Indeed %>%
  mutate(
    Date     = mdy(Date),
    Platform = "Indeed"
  ) %>%
  rename(ID = ...1)

# Combined
common_columns <- c("ID", "Title", "Position", "Current_employee", "Date", "Review", "Platform")

reviews <- bind_rows(
  Glassdoor %>% select(all_of(common_columns)),
  Indeed    %>% select(all_of(common_columns))
)

write.csv(reviews, "Glassdoor+Indeed.csv", row.names = FALSE)
```

### Tokenisation & Stop-Word Removal

```{r sa-tokenise, eval=TRUE}
tidy_reviews <- reviews %>%
  unnest_tokens(word, Review)

data("stop_words")

tidy_reviews <- tidy_reviews %>%
  anti_join(stop_words)
```

### Time Segmentation

Reviews are grouped into four meaningful periods relative to the analysis date.

```{r sa-time, eval=TRUE}
tidy_reviews <- tidy_reviews %>%
  mutate(
    time_segment = case_when(
      Date < as.Date("2020-06-01")                                     ~ "Pre-COVID",
      Date >= as.Date("2020-06-01") & Date < as.Date("2023-09-01")      ~ "Post-COVID to 12+ months ago (September 2023)",
      Date >= as.Date("2023-09-01")  & Date <as.Date("2024-02-01")        ~ "6-12 months ago (September 2023 - February 2024)",
      Date >= as.Date("2024-02-01")                                       ~ "Last 6 months (February 2024 - September 2024)",
      TRUE ~ NA_character_
    ),
    time_segment = factor(
      time_segment,
      levels = c("Last 6 months (February 2024 - September 2024)", "6-12 months ago (September 2023 - February 2024)", "Post-COVID to 12+ months ago (September 2023)", "Pre-COVID")
    )
  )
```

### Load Sentiment Lexicons

Three complementary lexicons are used to capture different dimensions of sentiment:

| Lexicon | What it measures |
|---------|-----------------|
| **NRC** | 8 discrete emotions + positive/negative |
| **Bing** | Binary positive vs. negative |
| **AFINN** | Numeric score from −5 (most negative) to +5 (most positive) |

```{r sa-lexicons, eval=TRUE}
sentiments         <- get_sentiments("nrc")
sentiments_PN      <- get_sentiments("bing")
sentiments_numeric <- get_sentiments("afinn")
```

### Join Reviews with Sentiments

```{r sa-join, eval=TRUE}
review_sentiments         <- tidy_reviews %>% inner_join(sentiments,         relationship = "many-to-many")
review_sentiments_PN      <- tidy_reviews %>% inner_join(sentiments_PN,      relationship = "many-to-many")
review_sentiments_numeric <- tidy_reviews %>% inner_join(sentiments_numeric, relationship = "many-to-many")

# Categorise AFINN scores into three buckets
review_sentiments_numeric <- review_sentiments_numeric %>%
  mutate(
    sentiment_category = case_when(
      value >= -5 & value <= -2 ~ "Negative",
      value >= -1 & value <= 1  ~ "Neutral",
      value >= 2  & value <= 5  ~ "Positive"
    )
  )
```

---

### Overall Sentiment Distribution (NRC)

```{r sa-nrc-overall, eval=TRUE}
sentiment_n <- review_sentiments %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment, sort = TRUE) %>%
  ungroup()

sentiment_n
```

**Key finding:** Across both platforms, the dominant emotion is *positive*, followed by *trust* and *anticipation* - consistent with Morgan Stanley's strong employer brand. Negative emotions (fear, anger, disgust) are present but in the minority.

The tables below show the full NRC emotion counts for each platform (former employees only):

**Indeed**

| Sentiment    | Count  |
|:-------------|-------:|
| positive     | 9,701  |
| trust        | 6,066  |
| anticipation | 3,647  |
| joy          | 2,535  |
| negative     | 2,078  |
| fear         |   995  |
| surprise     |   910  |
| sadness      |   884  |
| anger        |   799  |
| disgust      |   445  |

**Glassdoor**

| Sentiment    | Count  |
|:-------------|-------:|
| positive     | 5,245  |
| trust        | 3,198  |
| anticipation | 2,090  |
| joy          | 1,522  |
| negative     | 1,500  |
| sadness      |   690  |
| fear         |   679  |
| anger        |   672  |
| surprise     |   533  |
| disgust      |   414  |

Notably, the ordering of negative emotions differs between platforms: on Indeed, *fear* edges out *sadness*, while on Glassdoor the two are effectively tied - a subtle but consistent signal that Indeed reviewers frame dissatisfaction through job-security concerns, while Glassdoor reviewers lean more toward disappointment.

---

### Top Words per NRC Sentiment

```{r sa-nrc-words, eval=TRUE}
top_sentiments_former <- review_sentiments %>%
  filter(Current_employee == FALSE) %>%
  group_by(sentiment) %>%
  count(sentiment, word, sort = TRUE) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(.r = -row_number())

top_sentiments_former %>%
  ggplot(aes(.r, n, fill = sentiment)) +
  scale_x_continuous(
    breaks = top_sentiments_former$.r,
    labels = top_sentiments_former$word
  ) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ sentiment, nrow = 5, scales = "free") +
  labs(title = "Former Employees - Top Words per NRC Sentiment") +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(),
    axis.title.y       = element_blank()
  )

#ggsave("Top words per sentiment.tiff", height = 10, width = 6, dpi = 700)
```

---

### Positive vs. Negative (Bing)

```{r sa-bing-overall, eval=TRUE}
sentiment_n_PN <- review_sentiments_PN %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment, sort = TRUE) %>%
  ungroup()

sentiment_n_PN
```

```{r sa-bing-words, eval=TRUE}
top_sentiments_former_PN <- review_sentiments_PN %>%
  filter(Current_employee == FALSE) %>%
  group_by(sentiment) %>%
  count(sentiment, word, sort = TRUE) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(.r = -row_number())

top_sentiments_former_PN %>%
  ggplot(aes(.r, n, fill = sentiment)) +
  scale_x_continuous(
    breaks = top_sentiments_former_PN$.r,
    labels = top_sentiments_former_PN$word
  ) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ sentiment, nrow = 5, scales = "free") +
  labs(title = "Former Employees - Top Words: Positive vs. Negative (Bing)") +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y     = element_blank()
  )

#ggsave("Negative vs positive top words per sentiment.tiff", height = 10, width = 6, dpi = 700)
```

**Key finding:** The binary Bing lexicon confirms the NRC picture. On **Indeed**, positive words outnumber negative ones by more than 2:1 - **5,200 positive** vs. **2,345 negative**. On **Glassdoor** the margin is narrower but still clearly positive: **2,917 positive** vs. **1,824 negative**. The top positive drivers across both platforms cluster around words like *opportunity*, *professional*, and *smart*, while the top negative words centre on *hard* (often ambiguous - reflecting workload rather than purely pejorative use), *competitive*, and *bureaucratic*.

---

### Negative / Neutral / Positive (AFINN)

```{r sa-afinn-overall, eval=TRUE}
sentiment_n_numeric <- review_sentiments_numeric %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment_category, sort = TRUE) %>%
  ungroup()

sentiment_n_numeric
```

```{r sa-afinn-words, eval=TRUE}
top_words_afinn <- review_sentiments_numeric %>%
  filter(!is.na(sentiment_category)) %>%
  group_by(sentiment_category) %>%
  count(sentiment_category, word, sort = TRUE) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(.r = -row_number())

top_words_afinn %>%
  ggplot(aes(.r, n, fill = sentiment_category)) +
  scale_x_continuous(
    breaks = top_words_afinn$.r,
    labels = top_words_afinn$word
  ) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ sentiment_category, nrow = 3, scales = "free") +
  labs(title = "Former Employees - Top Words by AFINN Category") +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y     = element_blank()
  )

#ggsave("Pos-neut-neg top words per sentiment.tiff", height = 10, width = 6, dpi = 700)
```

**Key finding:** AFINN's three-way split adds resolution that Bing's binary categorisation misses. On **Indeed**, **4,076** tokens score as positive, **1,726** as neutral, and only **912** as negative - a healthy positive skew. **Glassdoor** shows the same ordering but compressed: **2,299 positive**, **1,342 neutral**, **836 negative**. The lower positive-to-negative ratio on Glassdoor (2.7×) compared with Indeed (4.5×) suggests that Glassdoor reviewers, despite their broadly positive framing, surface more granular criticism - in line with the platform's structured pros/cons format.

---

### Sentiment Trends Over Time

#### NRC Emotions by Time Period

```{r sa-nrc-time, eval=TRUE}
sentiment_n_by_time <- review_sentiments %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment, time_segment, sort = TRUE) %>%
  ungroup()

sentiment_n_by_time_by_percent <- sentiment_n_by_time %>%
  group_by(time_segment) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup()

sentiment_n_by_time_by_percent %>%
  mutate(time_segment = fct_relevel(
    time_segment, "Pre-COVID", "Post-COVID to 12+ months ago (September 2023)", "6-12 months ago (September 2023 - February 2024", "Last 6 months (February 2024 - September 2024"
  )) %>%
  ggplot(aes(x = sentiment, y = percent, fill = sentiment)) +
  geom_col() +
  facet_wrap(~ time_segment, ncol = 1) +
  labs(
    title = "NRC Emotion Distribution by Time Period",
    y = "Percentage (%)", x = NULL
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position    = "none",
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line()
  )

#ggsave("Top words per sentiment by time.tiff", height = 14, width = 8, dpi = 700)
```

#### Bing Positive vs. Negative by Time Period

```{r sa-bing-time, eval=TRUE}
sentiment_n_PN_by_time <- review_sentiments_PN %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment, time_segment, sort = TRUE) %>%
  ungroup()

sentiment_n_PN_by_time_by_percent <- sentiment_n_PN_by_time %>%
  group_by(time_segment) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup()

sentiment_n_PN_by_time_by_percent %>%
  mutate(time_segment = fct_relevel(
    time_segment, "Pre-COVID", "Post-COVID to 12+ months ago (September 2023)", "6-12 months ago (September 2023 - February 2024", "Last 6 months (February 2024 - September 2024"
  )) %>%
  ggplot(aes(x = sentiment, y = percent, fill = sentiment)) +
  geom_col() +
  facet_wrap(~ time_segment, nrow = 4) +
  labs(
    title = "Positive vs. Negative Sentiment by Time Period (Bing)",
    y = "Percentage (%)", x = NULL
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position    = "none",
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line()
  )

#ggsave("Top words per sentiment pos vs neg by time.tiff", height = 10, width = 6, dpi = 700)
```

#### AFINN Category Shares by Time Period

```{r sa-afinn-time, eval=TRUE}
sentiment_n_numeric_by_time <- review_sentiments_numeric %>%
  filter(Current_employee == FALSE) %>%
  count(sentiment_category, time_segment, sort = TRUE) %>%
  ungroup()

sentiment_n_numeric_by_time_by_percent <- sentiment_n_numeric_by_time %>%
  group_by(time_segment) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup()

sentiment_n_numeric_by_time_by_percent %>%
  mutate(time_segment = fct_relevel(
    time_segment, "Pre-COVID", "Post-COVID to 12+ months ago (September 2023)", "6-12 months ago (September 2023 - February 2024", "Last 6 months (February 2024 - September 2024"
  )) %>%
  ggplot(aes(x = sentiment_category, y = percent, fill = sentiment_category)) +
  geom_col() +
  facet_wrap(~ time_segment, ncol = 1) +
  labs(
    title = "AFINN Sentiment Category by Time Period",
    x = "Sentiment Category", y = "Percent (%)"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position    = "none",
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line()
  )

#ggsave("Top words per sentiment pos neutr neg by time.tiff", height = 10, width = 6, dpi = 700)
```

#### Continuous Sentiment Trend (AFINN Score over Date)

```{r sa-trend-facet, eval=TRUE}
review_sentiments_numeric %>%
  mutate(time_segment = fct_relevel(
    time_segment, "Pre-COVID", "Post-COVID to 12+ months ago (September 2023)", "6-12 months ago (September 2023 - February 2024", "Last 6 months (February 2024 - September 2024"
  )) %>%
  ggplot(aes(x = Date, y = value)) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ time_segment, scales = "free_x", ncol = 2, labeller = label_wrap_gen(width = 30)) +
  labs(
    title = "Average AFINN Sentiment Value Over Time (by Period)",
    x = "Date", y = "Sentiment Value"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(),
    strip.text         = element_text(size = 12, face = "bold")
  )

#ggsave("Average positivity of sentiment over time.tiff", height = 8, width = 10, dpi = 700)
```

```{r sa-trend-overall, eval=TRUE}
review_sentiments_numeric %>%
  ggplot(aes(x = Date, y = value)) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Overall AFINN Sentiment Value Over Time",
    x = "Date", y = "Sentiment Value"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line()
  )

#ggsave("Average positivity of sentiment over time without segregation.tiff", height = 8, width = 10, dpi = 700)
```

**Key finding - temporal trends:** All three lexicons point in the same direction across both platforms: **sentiment has declined from the pre-COVID era to the present**.

- On **Indeed**, the LOESS trend line starts declining around 2015–2016 and recently crosses into **negative territory** in the most recent window - the starkest signal in the dataset. Emotions of *fear* and *anger* have grown relative to *joy* across successive periods.
- On **Glassdoor**, the trajectory is similar though slightly less severe: sentiment has fallen consistently since 2020, but the **last 6 months (February–September 2024)** show a partial recovery toward *neutral* rather than tipping negative - a tentative sign of stabilisation.
- The **combined dataset** reinforces both signals: AFINN scores peaked around **2018** before entering a sustained post-COVID decline, with the merged LOESS curve sitting below pre-COVID baselines throughout the most recent period.

---

## Key Findings

### Indeed Reviews

- The dominant NRC emotion is **positive** (**9,701** tokens), followed by *trust* (6,066), *anticipation* (3,647), and *joy* (2,535). Negative emotions are in the minority, led by *fear* (995) and *surprise* (910), with *disgust* the least represented (445).
- The positive framing holds across all three lexicons: Bing returns **5,200 positive** vs. **2,345 negative** words; AFINN shows **4,076 positive**, **1,726 neutral**, and only **912 negative** tokens - a positive-to-negative ratio of roughly 4.5:1.
- Temporally, the average sentiment has **declined from the pre-COVID period to the present**, with *fear* and *anger* growing and *joy* receding across successive periods. The AFINN trend line, which peaked around 2015–2016, crosses into **negative territory** in the most recent 6-month window - the most striking single finding in the dataset.

### Glassdoor Reviews

- Results mirror Indeed: *positive* leads (**5,245**), followed by *trust* (**3,198**) and *anticipation* (**2,090**). Unlike Indeed, *sadness* ranks above *fear* on Glassdoor - reflecting the platform's structured pros/cons format, which may elicit more reflective, disappointment-oriented language.
- Bing: **2,917 positive** vs. **1,824 negative** (ratio ≈ 1.6:1 - lower than Indeed's 2.2:1). AFINN: **2,299 positive**, **1,342 neutral**, **836 negative**.
- The downward trend is visible here too, but the **last 6-month window shows a partial recovery toward neutral** rather than negative - a tentative sign of stabilisation not present in the Indeed data.

### Combined Picture

When both datasets are merged, the long-run AFINN trend confirms a **peak around 2018** followed by a sustained decline through the post-COVID period. The merged LOESS curve sits below pre-COVID baselines throughout the most recent window, suggesting structural - rather than event-driven - shifts in employee experience at the firm.
